---
title: "Simulating species abundances"
author: "Andy Rominger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating species abundances}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))


```{r, eval = FALSE}
library(roleLite)
library(pika)

S <- 500
aa <- matrix(runif(S^2, 0, 5), nrow = S)
diag(aa) <- runif(S, 1, 10)

# aa <- matrix(1, nrow = S, ncol = S)

x <- rlseries(S, 0.001)

boo <- compSim(J = 1000, nu = 0, m = 0.01, alpha = aa,
               niter = 100000, niterTimestep = 100,
               initMeta = x)

# boo <- untbSim(J = 1000, nu = 0, m = 0.1,
#                niter = 10000, niterTimestep = 100,
#                initMeta = x)

doo <- getAbund(boo)


plot(doo$tstep, doo$abund, type = 'n', log = 'y')
for(i in 1:S) {
    lines(doo$tstep[doo$spID == i],
          doo$abund[doo$spID == i],
          col = viridis::viridis(S)[i])
}

plot(sad(getAbund(boo, nrow(boo))$abund), ptype = 'rad', log = 'y')
```


